---
title: 'Graphs from Jennifer'
author: "Jen"
output: 
  html_document:
    keep_md: TRUE
    toc: TRUE
    toc_float: TRUE
    df_print: paged
    code_download: true
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)     # for data cleaning and plotting
library(lubridate)     # for date manipulation
library(openintro)     # for the abbr2state() function
library(maps)          # for map data
library(ggmap)         # for mapping points on maps
library(gplots)        # for col2hex() function
library(RColorBrewer)  # for color palettes
library(sf)            # for working with spatial data
library(leaflet)       # for highly customizable mapping
library(ggthemes)      # for more themes (including theme_map())
library(plotly)        # for the ggplotly() - basic interactivity
library(gganimate)     # for adding animation layers to ggplots
library(transformr)    # for "tweening" (gganimate)
library(gifski)        # need the library for creating gifs but don't need to load each time
library(shiny)         # for creating interactive apps
library(ggimage)
library(tidytuesdayR)
theme_set(theme_minimal())
```

```{r}
# College Tuition Data
tuition_cost <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-10/tuition_cost.csv')

tuition_income <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-10/tuition_income.csv') 

salary_potential <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-10/salary_potential.csv')

historical_tuition <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-10/historical_tuition.csv')

diversity_school <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-10/diversity_school.csv')
```

```{r}
# Wealth and Income Over Time
lifetime_earn <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/lifetime_earn.csv')
student_debt <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/student_debt.csv')
retirement <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/retirement.csv')
home_owner <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/home_owner.csv')
race_wealth <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/race_wealth.csv')
income_time <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/income_time.csv')
income_limits <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/income_limits.csv')
income_aggregate <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/income_aggregate.csv')
income_distribution <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/income_distribution.csv')
income_mean <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/income_mean.csv')
```



```{r}
student_debt %>%  
  left_join(retirement, by =c("race", "year")) %>% 
  ggplot() +
  geom_line(aes(x = loan_debt,
             y = retirement,
             color = race)) + 
  scale_color_manual(values = c("#40E0D0","#6495ED", "#9370DB")) + 
  labs(title = "Average Retirement Savings Based on Student Loan Debt",
       x = "Loan Debt",
       y = "Retirement Savings",
       color = "Race") 
```

* Referring to Kate's patchwork graphs on student loan debt and enrollment based on racial groups, this graph shows that even though White students have a higher enrollment rate and have roughly the same amount of student loan debt, they nonetheless have significantly higher retirement savings. This is likely due to the wage gap in racial groups and the employment rate for people of color in higher paying positions. 


```{r}
income_limits %>%  
  left_join(income_time, by =c("year")) %>% 
  filter(race %in% c("White Alone", "Black Alone", "Hispanic")) %>% 
  mutate(ordered = fct_reorder(income_quintile,
                               income_dollars,
                               .desc = FALSE)) %>% 
  ggplot() +
  geom_boxplot(aes(x = ordered,
             y = income_dollars,
             fill = race)) + 
  scale_y_continuous(limits = c(0,200000), expand = c(0, 0),labels=scales::dollar_format()) + 
  labs(title = "Income limits based on racial groups",
       x = "",
       y = "",
       fill = "Race") + 
  scale_fill_manual(values = c("#40E0D0","#6495ED", "#9370DB"))
```



